A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-10 tex/sec1-intro.tex
[ tex ] Explain decidability a bit further in S1

Our work is partly inspired by \varcitet{Wadler2022}{'s} textbook, where programming language concepts are formally introduced using the proof assistant~\Agda based on Martin-L\"of Type Theory, and one particular topic is bidirectional type synthesis for~\PCF.
Their treatment deviates from existing literature:
Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
By contrast, \citeauthor{Wadler2022} exploit the simultaneously computational and logical nature of Martin-L\"of type theory to formulate algorithmic soundness, completeness, and decidability in one go.
%(although they do not emphasise the difference of their approach from the traditional one)
Recall that the law of excluded middle $P + \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, is a way to compare arbitrary $x$~and~$y$ and decide whether they are equal, which may or may not be possible depending on the domain of $x$~and~$y$.
Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ has a derivation for some~$A$'.
\end{quote}
%echoing \citeauthor{Wadler2022}'s formulation as noted in \cref{re:type-synthesis-as-decidability-proof}.
Then, both algorithmic soundness and completeness are implied.
%the proof either yields a typing derivation for the given raw term~$t$, or a proof that such a derivation is impossible; the former case is algorithmic soundness, and the latter is algorithmic completeness in contrapositive form.
%\todo{traditional completeness is still derivable from this `LEM'}
Our work takes the same approach but expands the scope significantly.
%That is, our decidable bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.

>>>>

Our work is partly inspired by \varcitet{Wadler2022}{'s} textbook, where programming language concepts are formally introduced using the proof assistant~\Agda based on Martin-L\"of Type Theory, and one particular topic is bidirectional type synthesis for~\PCF.
Their treatment deviates from existing literature:
Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
By contrast, \citeauthor{Wadler2022} exploit the simultaneously computational and logical nature of Martin-L\"of type theory to formulate algorithmic soundness, completeness, and decidability in one go.
%(although they do not emphasise the difference of their approach from the traditional one)
Recall that the law of excluded middle $P + \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether they are equal and gives an identity proof explicitly if they are, which may or may not be possible depending on the domain of $x$~and~$y$.
Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ has a derivation for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a contradiction proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Then, both algorithmic soundness and completeness in original form are implied.
%echoing \citeauthor{Wadler2022}'s formulation as noted in \cref{re:type-synthesis-as-decidability-proof}.
%\todo{traditional completeness is still derivable from this `LEM'}
Our work takes the same approach but expands the scope significantly.
%That is, our decidable bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.
